---
title: "PA3"
author: "Bernardo Lopez, Victor Zahn, Gregory Morita, Jonathan Garcia"
format: html
editor: visual
---

```{r}
library(tidyverse)
library(knitr)
library(kableExtra)
library(gt)
```

## Dataset

Two datasets:

Extreme poverty rate (\$2.15/day, Gapminder data) long series 1800 to 2100

```{r}

poverty_unclean <- read.csv("./Data/gm_epov_rate.csv")

```

Life expectancy (The number of years a newborn infant would live if the current mortality rates at different ages were to stay the same throughout its life.)

```{r}

expectancy_unclean <- read.csv("./Data/lex.csv")

```

Data Cleaning

```{r}

poverty <- poverty_unclean |>
  pivot_longer(cols = "X1800":"X2100",
               names_to = "year",
               values_to = "poverty_rate") |>
  filter(!is.na(poverty_rate)) |> # Gets rid of rows with no data
  mutate(year = as.numeric(str_extract(year, pattern = "[:digit:]{4}")), # Extract years
         poverty_rate = poverty_rate/100) # Convert rates to range 0-1

expectancy <- expectancy_unclean |>
  pivot_longer(cols = "X1800":"X2100",
               names_to = "year",
               values_to = "life_expectancy") |>
  filter(!is.na(life_expectancy) & life_expectancy > 0) |> # Get rid of rows with no data (and 0 seems extreme)
  mutate(year = as.numeric(str_extract(year, pattern = "[:digit:]{4}"))) # Extract years
```

Data Joining

```{r}
combined <- poverty |>
  full_join(expectancy, join_by(country == country, year == year)) |>
  drop_na()

```

## Written Analysis

**Hypothesis**

Global modernization and advances in healthcare have contributed to increases in life expectancy in the modern era. These benefits are often unevenly distributed and tend to be concentrated among populations with higher socioeconomic status(Liu et al., 2024). In this study, we will analyze historical poverty rates and life expectancy data to examine if extreme poverty rates have a linear negative relationship with life expectancy rates.

The data cleaning should have little impact on observations, though it does include removing any NA values, which may obscure the overall picture for any one nation. Missing values are typically in large chunks, and shouldn't impact trends in nations that do have data.

**References**

Lili Liu et al. (2024), Impacts of Poverty and Lifestyles on Mortality: A cohort Study in Predominantly Low-Income Americans, American Journal of Preventive Medicine.67(1)15-23.
